\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2024-01-01 misc/JSS article/jss-article-rnw/article.tex
Adding JSS article.

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, cache=TRUE}
% \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}

<<preliminaries, echo=FALSE, message=FALSE, warning=FALSE, results=hide>>=
# knitr::opts_chunk$set(cache = FALSE, size = "small")
options(prompt = "R> ", continue = "+  ", width = 80, digits = 4, useFancyQuotes = FALSE, warn = 1)
library(data.table)
library(magrittr)
library(microbenchmark)
library(collapse)
@

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g. \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents.} \newline

@@

<<collapse_topics>>=
.COLLAPSE_TOPICS
help("collapse-documentation")
@

@@

<<aq>>=
fnobs(airquality)
@

@@

<<aq_imp>>=
fmedian(airquality[1:2], airquality$Month, TRA = "replace_na", set = TRUE)
@

@@

<<aq_trans>>=
airquality |> fmutate(
  rad_day = fsum(as.double(Solar.R), Day, TRA = "/"),
  ozone_deg = Ozone / Temp,
  ozone_amed = Ozone > fmedian(Ozone, Month, TRA = "fill"),
  ozone_resid = fmean(Ozone, list(Month, ozone_amed), ozone_deg, "-")
) |> head(3)
@

@@

<<GRP>>=
str(g <- GRP(mtcars, ~ cyl + vs + am))
@

@@

<<gcomp>>=
dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt
add_vars(g$groups,
  fmean(dat, g, w, use.g.names = FALSE) |> add_stub("w_mean_"),
  fsd(dat, g, w, use.g.names = FALSE) |> add_stub("w_sd_")) |> head(2)
@

@@

<<gtrans>>=
mtcars |> add_vars(fmean(dat, g, w, "-") |> add_stub("w_demean_"),
                   fscale(dat, g, w) |> add_stub("w_scale_")) |> head(2)
@

@@

<<fsummarise>>=
mtcars |>
  fsubset(mpg > 11) |>
  fgroup_by(cyl, vs, am) |>
  fsummarise(across(c(mpg, carb, hp), fmean),
             qsec_w_med = fmean(qsec, wt)) |> head(2)
@

@@

<<fsummarise_2>>=
mtcars |>
  fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
  fgroup_by(cyl, vs, am) |>
  fmean(wt) |> head(2)
@

@@

<<flinreg>>=
mtcars |>
 fgroup_by(vs) |>
 fmutate(dm_carb = fmean(carb, TRA = "-")) |>
 fsummarise(slope = fsum(mpg, dm_carb) %/=% fsum(dm_carb^2))
@

@@

<<wstats>>=
mtcars |>
    fgroup_by(cyl, vs, am) |>
    fmutate(o = radixorder(GRPid(), mpg)) |>
    fsummarise(mpg_min = fmin(mpg),
               mpg_Q1 = fnth(mpg, 0.25, wt, o = o, ties = "q8"),
               mpg_mean = fmean(mpg, wt),
               mpg_median = fmedian(mpg, wt, o = o, ties = "q8"),
               mpg_mode = fmode(mpg, wt, ties = "max"),
               mpg_Q3 = fnth(mpg, 0.75, wt, o = o, ties = "q8"),
               mpg_max = fmax(mpg)) |> head(3)
@

@@

<<collap>>=
collap(wlddev, country + PCGDP + LIFEEX ~ year + income, w = ~ POP) |>
  head(4)
@

@@

<<ts_example>>=
fgrowth(airmiles) |> round(2)
@

@@

<<ts_example_ctd>>=
am_ir <- airmiles[-c(3, 15)]
t <- time(airmiles)[-c(3, 15)]
fgrowth(am_ir, t = t) |> round(2)
fgrowth(am_ir, -1:3, t = t) |> head(4)
@

@@

<<example_growth>>=
G(wlddev, c(1, 10), by = POP + LIFEEX ~ iso3c, t = ~ year) |> head(3)
settransform(wlddev, POP_growth = G(POP, g = iso3c, t = year))
@

@@

<<example_growth_continued>>=
wlddev |> fgroup_by(iso3c) |> fselect(iso3c, year, POP, LIFEEX) |>
  fmutate(across(c(POP, LIFEEX), G, t = year)) |> head(2)
@

@@

<<example_scale_cener>>=
iris |> fgroup_by(Species) |> fscale() |> head(2)
STD(iris, ~ Species) |> head(2)
@

@@

<<example_reg>>=
lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
@

@@

<<example_HD>>=
HDW(wlddev, PCGDP + LIFEEX ~ iso3c * poly(year, 3), stub = F) |> head(2)
@

@@

<<indexing>>=
wldi <- wlddev |> findex_by(iso3c, year)
wldi |> fsubset(-3, iso3c, year, PCGDP:POP) |> G() |> head(4)
@
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties.
<<indexing_2>>=
LIFEEXi = wldi$LIFEEX
str(LIFEEXi, width = 70, strict = "cut")
c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5]))
G(LIFEEXi[c(1:5, 7:10)])
@
The transformation and estimation below demonstrate the deep indexation of \\ \class{indexed\_frame}'s, allowing correct computations in arbitrary data masking environments.
<<indexing_3>>=
settransform(wldi, PCGDP_ld = Dlog(PCGDP))
lm(D(LIFEEX) ~ L(PCGDP_ld, 0:5) + B(PCGDP_ld), wldi) |>
  summary() |> coef() |> round(3)
@
The above example could also have been executed in one line as \code{lm(D(LIFEEX) ~ } \\ \code{L(Dlog(PCGDP), 0:5) + B(Dlog(PCGDP)), wldi)}, log-differencing \code{PCGDP} twice. \newline

In comparison with existing solutions, the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.}. Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures for time series operations or native handling of irregularity in basic operations. \class{indexed\_series} and \class{indexed\_frame}, on the other hand, work anywhere and can be superimposed on any suitable object (such as \class{sf} to create a spatiotemporal panel), as long as \pkg{collapse}'s functions (\code{flag()/L()} etc.) are used to perform the time-based computations. The \class{index\_df} attached to these objects can be used with other general tools such as \code{collapse::BY()} to perform grouped computations on these objects with 3rd-party functions. An example of calculating a 5-year rolling average is given below. Last but not least, the performance of these classes is second to none, as demonstrated in the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{on slide 40}.
%
<<rollmean>>=
BY(LIFEEXi, findex(LIFEEXi)$iso3c, data.table::frollmean, 5) |> head(10)
@
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
While \pkg{collapse} has a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{Data Manipulation Functions}, its implementations of table joins and pivots is particularly noteworthy since they offer several new features such as various verbose options for table joins, pivots supporting variable labels, and 'recast' pivots.
%
\subsection{Joins}
%
Compared to commercial environments such as \proglang{STATA}, the implementation of joins in most open-source software, including \proglang{R}, is non-verbose, i.e., provides no information on how many and which records were joined from both tables. This is somewhat unsatisfying and often provokes manual efforts to validate the join operation. \code{collapse::join} provides a rich set of options to make table join operations intelligible. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
     multiple = FALSE, sort = FALSE, keep.col.order = TRUE,
     drop.dup.cols = FALSE, verbose = 1, column = NULL, attr = NULL, ...)
\end{Code}
By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. Users can request the generation of a '.join' column, akin to \proglang{STATA}'s '\_merge' column, indicating the origin of records in the joined table, as shown below.
%
<<joins>>=
df1 <- data.frame(id1 = c(1, 1, 2, 3),
                  id2 = c("a", "b", "b", "c"),
                  name = c("John", "Jane", "Bob", "Carl"),
                  age = c(35, 28, 42, 50))
df2 <- data.frame(id1 = c(1, 2, 3, 3),
                  id2 = c("a", "b", "c", "e"),
                  salary = c(60000, 55000, 70000, 80000),
                  dept = c("IT", "Marketing", "Sales", "IT"))

join(df1, df2, on = c("id1", "id2"), how = "full", column = TRUE)
@
%
An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}). Users can also invoke the \code{validate} argument to check the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
%
<<joins_2>>=
join(df1, df2, on = c("id1", "id2"), validate = "1:1", attr = "join") |>
  attr("join") |> str(width = 70, strict = "cut")
@
%
A few further particularities are worth highlighting. First, \code{collapse::join} is also class-agnostic and preserves the attributes of \code{x}. It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"} or \code{"anti"}) and default to \code{"left"}, so the default behavior simply adds columns to \code{x}. By default (\code{sort = FALSE}), the order of rows in \code{x} is also preserved. Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys. } Additionally, by default (\code{multiple = FALSE}), only the first matches from both tables are joined to avoid silent cartesian duplication of records. In multi-match settings, this will be reflected by few records from \code{y} being used. \fct{fmatch} also has a built-in overidentification check, which issues a warning if more key columns than necessary to identify the records are provided.
%
<<join_3>>=
df2$name = df1$name
join(df1, df2) |> capture.output(type="m") |> strwrap(77) |> cat(sep="\n")
@
%
% This warning can be silenced by passing \code{overid = 2} or restricting the join columns using \code{on}.
A final noteworthy feature is the handling of duplicate non-id columns in both tables.
%
<<join_4>>=
join(df1, df2, on = c("id1", "id2"))
@
%
By default (\code{suffix = NULL}), \fct{join} extracts the name of the \code{y} table and appends \code{y}-columns with it. \code{x}-columns are not renamed. This is congruent to the principle of adding columns to \code{x} and altering this table as little as possible. Another option, \code{drop.dup.cols = "x"/"y"} can be used to simply drop duplicate columns from \code{x} or \code{y} before the join operation.
%
\subsection{Pivots}
%
The reshaping/pivoting functionality of commercial and open source software is also unsatisfying for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data. Such datasets can only be reshaped by losing these labels or additional manual efforts to retain them. Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/ tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both. Finally, since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is no modern replacement for \code{reshape2::recast()}. This requires \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of syntax and memory for such frequently required data transpositions. \newline

>>>>

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}

% \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++ based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g. \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents.} \newline

@@

\begin{Schunk}
\begin{Sinput}
R> .COLLAPSE_TOPICS
\end{Sinput}
\begin{Soutput}
 [1] "collapse-documentation"     "fast-statistical-functions"
 [3] "fast-grouping-ordering"     "fast-data-manipulation"    
 [5] "quick-conversion"           "advanced-aggregation"      
 [7] "data-transformations"       "time-series-panel-series"  
 [9] "list-processing"            "summary-statistics"        
[11] "recode-replace"             "efficient-programming"     
[13] "small-helpers"              "collapse-options"          
\end{Soutput}
\begin{Sinput}
R> help("collapse-documentation")
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> fnobs(airquality)
\end{Sinput}
\begin{Soutput}
  Ozone Solar.R    Wind    Temp   Month     Day 
    116     146     153     153     153     153 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> fmedian(airquality[1:2], airquality$Month, TRA = "replace_na", set = TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> airquality |> fmutate(
+    rad_day = fsum(as.double(Solar.R), Day, TRA = "/"),
+    ozone_deg = Ozone / Temp,
+    ozone_amed = Ozone > fmedian(Ozone, Month, TRA = "fill"),
+    ozone_resid = fmean(Ozone, list(Month, ozone_amed), ozone_deg, "-")
+  ) |> head(3)
\end{Sinput}
\begin{Soutput}
  Ozone Solar.R Wind Temp Month Day rad_day ozone_deg ozone_amed ozone_resid
1    41     190  7.4   67     5   1   0.191    0.6119       TRUE     -10.279
2    36     118  8.0   72     5   2   0.135    0.5000       TRUE     -15.279
3    12     149 12.6   74     5   3   0.168    0.1622      FALSE      -3.035
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> str(g <- GRP(mtcars, ~ cyl + vs + am))
\end{Sinput}
\begin{Soutput}
Class 'GRP'  hidden list of 9
 $ N.groups    : int 7
 $ group.id    : int [1:32] 4 4 3 5 6 5 6 2 2 5 ...
 $ group.sizes : int [1:7] 1 3 7 3 4 12 2
 $ groups      :'data.frame':	7 obs. of  3 variables:
  ..$ cyl: num [1:7] 4 4 4 6 6 8 8
  ..$ vs : num [1:7] 0 1 1 0 1 0 0
  ..$ am : num [1:7] 1 0 1 1 0 0 1
 $ group.vars  : chr [1:3] "cyl" "vs" "am"
 $ ordered     : Named logi [1:2] TRUE FALSE
  ..- attr(*, "names")= chr [1:2] "ordered" "sorted"
 $ order       : int [1:32] 27 8 9 21 3 18 19 20 26 28 ...
  ..- attr(*, "starts")= int [1:7] 1 2 5 12 15 19 31
  ..- attr(*, "maxgrpn")= int 12
  ..- attr(*, "sorted")= logi FALSE
 $ group.starts: int [1:7] 27 8 3 1 4 5 29
 $ call        : language GRP.default(X = mtcars, by = ~cyl + vs + am)
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt
R> add_vars(g$groups,
+    fmean(dat, g, w, use.g.names = FALSE) |> add_stub("w_mean_"),
+    fsd(dat, g, w, use.g.names = FALSE) |> add_stub("w_sd_")) |> head(2)
\end{Sinput}
\begin{Soutput}
  cyl vs am w_mean_mpg w_mean_disp w_sd_mpg w_sd_disp
1   4  0  1      26.00       120.3    0.000       0.0
2   4  1  0      23.02       137.1    1.236      11.6
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> mtcars |> add_vars(fmean(dat, g, w, "-") |> add_stub("w_demean_"),
+                     fscale(dat, g, w) |> add_stub("w_scale_")) |> head(2)
\end{Sinput}
\begin{Soutput}
              mpg cyl disp  hp drat    wt  qsec vs am gear carb w_demean_mpg
Mazda RX4      21   6  160 110  3.9 2.620 16.46  0  1    4    4       0.4357
Mazda RX4 Wag  21   6  160 110  3.9 2.875 17.02  0  1    4    4       0.4357
              w_demean_disp w_scale_mpg w_scale_disp
Mazda RX4             5.027      0.6657       0.6657
Mazda RX4 Wag         5.027      0.6657       0.6657
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> mtcars |>
+    fsubset(mpg > 11) |>
+    fgroup_by(cyl, vs, am) |>
+    fsummarise(across(c(mpg, carb, hp), fmean),
+               qsec_w_med = fmean(qsec, wt)) |> head(2)
\end{Sinput}
\begin{Soutput}
  cyl vs am  mpg  carb    hp qsec_w_med
1   4  0  1 26.0 2.000 91.00      16.70
2   4  1  0 22.9 1.667 84.67      21.04
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> mtcars |>
+    fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
+    fgroup_by(cyl, vs, am) |>
+    fmean(wt) |> head(2)
\end{Sinput}
\begin{Soutput}
  cyl vs am sum.wt   mpg carb   hp
1   4  0  1  2.140 26.00 2.00 91.0
2   4  1  0  8.805 23.02 1.72 83.6
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> mtcars |>
+   fgroup_by(vs) |>
+   fmutate(dm_carb = fmean(carb, TRA = "-")) |>
+   fsummarise(slope = fsum(mpg, dm_carb) %/=% fsum(dm_carb^2))
\end{Sinput}
\begin{Soutput}
  vs   slope
1  0 -0.5557
2  1 -2.0706
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> mtcars |>
+      fgroup_by(cyl, vs, am) |>
+      fmutate(o = radixorder(GRPid(), mpg)) |>
+      fsummarise(mpg_min = fmin(mpg),
+                 mpg_Q1 = fnth(mpg, 0.25, wt, o = o, ties = "q8"),
+                 mpg_mean = fmean(mpg, wt),
+                 mpg_median = fmedian(mpg, wt, o = o, ties = "q8"),
+                 mpg_mode = fmode(mpg, wt, ties = "max"),
+                 mpg_Q3 = fnth(mpg, 0.75, wt, o = o, ties = "q8"),
+                 mpg_max = fmax(mpg)) |> head(3)
\end{Sinput}
\begin{Soutput}
  cyl vs am mpg_min mpg_Q1 mpg_mean mpg_median mpg_mode mpg_Q3 mpg_max
1   4  0  1    26.0  26.00    26.00      26.00     26.0  26.00    26.0
2   4  1  0    21.5  22.10    23.02      23.17     24.4  24.38    24.4
3   4  1  1    21.4  22.29    27.74      27.85     30.4  31.79    33.9
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> collap(wlddev, country + PCGDP + LIFEEX ~ year + income, w = ~ POP) |>
+    head(4)
\end{Sinput}
\begin{Soutput}
        country year              income   PCGDP LIFEEX       POP
1 United States 1960         High income 12768.7  68.59 7.495e+08
2      Ethiopia 1960          Low income   658.5  38.33 1.474e+08
3         India 1960 Lower middle income   500.8  45.27 9.280e+08
4         China 1960 Upper middle income  1166.1  49.86 1.184e+09
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> fgrowth(airmiles) |> round(2)
\end{Sinput}
\begin{Soutput}
Time Series:
Start = 1937 
End = 1960 
Frequency = 1 
 [1]    NA 16.50 42.29 54.03 31.65  2.38 15.23 33.29 54.36 76.92  2.71 -2.10
[13] 12.91 18.51 32.03 18.57 17.82 13.61 18.19 12.83 13.32  0.01 15.49  4.25
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> am_ir <- airmiles[-c(3, 15)]
R> t <- time(airmiles)[-c(3, 15)]
R> fgrowth(am_ir, t = t) |> round(2)
\end{Sinput}
\begin{Soutput}
 [1]    NA 16.50    NA 31.65  2.38 15.23 33.29 54.36 76.92  2.71 -2.10 12.91
[13] 18.51    NA 17.82 13.61 18.19 12.83 13.32  0.01 15.49  4.25
\end{Soutput}
\begin{Sinput}
R> fgrowth(am_ir, -1:3, t = t) |> head(4)
\end{Sinput}
\begin{Soutput}
         FG1   --    G1  L2G1  L3G1
[1,] -14.167  412    NA    NA    NA
[2,]      NA  480 16.50    NA    NA
[3,] -24.043 1052    NA 119.2 155.3
[4,]  -2.327 1385 31.65    NA 188.5
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> G(wlddev, c(1, 10), by = POP + LIFEEX ~ iso3c, t = ~ year) |> head(3)
\end{Sinput}
\begin{Soutput}
  iso3c year G1.POP L10G1.POP G1.LIFEEX L10G1.LIFEEX
1   AFG 1960     NA        NA        NA           NA
2   AFG 1961  1.917        NA     1.590           NA
3   AFG 1962  1.985        NA     1.544           NA
\end{Soutput}
\begin{Sinput}
R> settransform(wlddev, POP_growth = G(POP, g = iso3c, t = year))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> wlddev |> fgroup_by(iso3c) |> fselect(iso3c, year, POP, LIFEEX) |>
+    fmutate(across(c(POP, LIFEEX), G, t = year)) |> head(2)
\end{Sinput}
\begin{Soutput}
  iso3c year     POP LIFEEX G1.POP G1.LIFEEX
1   AFG 1960 8996973  32.45     NA        NA
2   AFG 1961 9169410  32.96  1.917      1.59
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> iris |> fgroup_by(Species) |> fscale() |> head(2)
\end{Sinput}
\begin{Soutput}
  Species Sepal.Length Sepal.Width Petal.Length Petal.Width
1  setosa       0.2667      0.1899       -0.357     -0.4365
2  setosa      -0.3007     -1.1291       -0.357     -0.4365
\end{Soutput}
\begin{Sinput}
R> STD(iris, ~ Species) |> head(2)
\end{Sinput}
\begin{Soutput}
  Species STD.Sepal.Length STD.Sepal.Width STD.Petal.Length STD.Petal.Width
1  setosa           0.2667          0.1899           -0.357         -0.4365
2  setosa          -0.3007         -1.1291           -0.357         -0.4365
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
\end{Sinput}
\begin{Soutput}
 (Intercept)         carb B(carb, cyl) 
     34.8297      -0.4655      -4.7750 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> HDW(wlddev, PCGDP + LIFEEX ~ iso3c * poly(year, 3), stub = F) |> head(2)
\end{Sinput}
\begin{Soutput}
   PCGDP   LIFEEX
1  9.964 0.023670
2 14.045 0.006743
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> wldi <- wlddev |> findex_by(iso3c, year)
R> wldi |> fsubset(-3, iso3c, year, PCGDP:POP) |> G() |> head(4)
\end{Sinput}
\begin{Soutput}
  iso3c year G1.PCGDP G1.LIFEEX G1.GINI G1.ODA G1.POP
1   AFG 1960       NA        NA      NA     NA     NA
2   AFG 1961       NA     1.590      NA  98.75  1.917
3   AFG 1963       NA        NA      NA     NA     NA
4   AFG 1964       NA     1.448      NA  24.48  2.112

Indexed by:  iso3c [1] | year [4 (61)] 
\end{Soutput}
\end{Schunk}
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties.
\begin{Schunk}
\begin{Sinput}
R> LIFEEXi = wldi$LIFEEX
R> str(LIFEEXi, width = 70, strict = "cut")
\end{Sinput}
\begin{Soutput}
 'indexed_series' num [1:13176] 32.4 33 33.5 34 34.5 ...
 - attr(*, "label")= chr "Life expectancy at birth, total (years)"
 - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
  ..$ iso3c: Factor w/ 216 levels "ABW","AFG","AGO",..: 2 2 2 2 2 2 ..
  .. ..- attr(*, "label")= chr "Country Code"
  ..$ year : Ord.factor w/ 61 levels "1960"<"1961"<..: 1 2 3 4 5 6 7..
  .. ..- attr(*, "label")= chr "Year"
\end{Soutput}
\begin{Sinput}
R> c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5]))
\end{Sinput}
\begin{Soutput}
[1] FALSE  TRUE
\end{Soutput}
\begin{Sinput}
R> G(LIFEEXi[c(1:5, 7:10)])
\end{Sinput}
\begin{Soutput}
[1]    NA 1.590 1.544 1.494 1.448    NA 1.366 1.362 1.365

Indexed by:  iso3c [1] | year [9 (61)] 
\end{Soutput}
\end{Schunk}
The transformation and estimation below demonstrate the deep indexation of \\ \class{indexed\_frame}'s, allowing correct computations in arbitrary data masking environments.
\begin{Schunk}
\begin{Sinput}
R> settransform(wldi, PCGDP_ld = Dlog(PCGDP))
R> lm(D(LIFEEX) ~ L(PCGDP_ld, 0:5) + B(PCGDP_ld), wldi) |>
+    summary() |> coef() |> round(3)
\end{Sinput}
\begin{Soutput}
                   Estimate Std. Error t value Pr(>|t|)
(Intercept)           0.299      0.007  44.412    0.000
L(PCGDP_ld, 0:5)--    0.300      0.080   3.735    0.000
L(PCGDP_ld, 0:5)L1    0.269      0.081   3.332    0.001
L(PCGDP_ld, 0:5)L2    0.227      0.079   2.854    0.004
L(PCGDP_ld, 0:5)L3    0.200      0.078   2.563    0.010
L(PCGDP_ld, 0:5)L4    0.143      0.076   1.871    0.061
L(PCGDP_ld, 0:5)L5    0.095      0.073   1.301    0.193
B(PCGDP_ld)          -1.021      0.316  -3.234    0.001
\end{Soutput}
\end{Schunk}
The above example could also have been executed in one line as \code{lm(D(LIFEEX) ~ } \\ \code{L(Dlog(PCGDP), 0:5) + B(Dlog(PCGDP)), wldi)}, log-differencing \code{PCGDP} twice. \newline

In comparison with existing solutions, the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.}. Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures for time series operations or native handling of irregularity in basic operations. \class{indexed\_series} and \class{indexed\_frame}, on the other hand, work anywhere and can be superimposed on any suitable object (such as \class{sf} to create a spatiotemporal panel), as long as \pkg{collapse}'s functions (\code{flag()/L()} etc.) are used to perform the time-based computations. The \class{index\_df} attached to these objects can be used with other general tools such as \code{collapse::BY()} to perform grouped computations on these objects with 3rd-party functions. An example of calculating a 5-year rolling average is given below. Last but not least, the performance of these classes is second to none, as demonstrated in the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{on slide 40}.
%
\begin{Schunk}
\begin{Sinput}
R> BY(LIFEEXi, findex(LIFEEXi)$iso3c, data.table::frollmean, 5) |> head(10)
\end{Sinput}
\begin{Soutput}
 [1]    NA    NA    NA    NA 33.46 33.96 34.46 34.95 35.43 35.92

Indexed by:  iso3c [1] | year [10 (61)] 
\end{Soutput}
\end{Schunk}
\section{Table joins and pivots} \label{sec:join_pivot}
While \pkg{collapse} has a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{Data Manipulation Functions}, its implementations of table joins and pivots is particularly noteworthy since they offer several new features such as various verbose options for table joins, pivots supporting variable labels, and 'recast' pivots.
\subsection{Joins}
Compared to commercial environments such as \proglang{STATA}, the implementation of joins in most open-source software, including \proglang{R}, is non-verbose, i.e., provides no information on how many and which records were joined from both tables. This is somewhat unsatisfying and often provokes manual efforts to validate the join operation. \code{collapse::join} provides a rich set of options to make table join operations intelligible. Its syntax is:
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
     multiple = FALSE, sort = FALSE, keep.col.order = TRUE,
     drop.dup.cols = FALSE, verbose = 1, column = NULL, attr = NULL, ...)
By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. Users can request the generation of a '.join' column, akin to \proglang{STATA}'s '\_merge' column, indicating the origin of records in the joined table, as shown below.
\begin{Schunk}
\begin{Sinput}
R> df1 <- data.frame(id1 = c(1, 1, 2, 3),
+                    id2 = c("a", "b", "b", "c"),
+                    name = c("John", "Jane", "Bob", "Carl"),
+                    age = c(35, 28, 42, 50))
R> df2 <- data.frame(id1 = c(1, 2, 3, 3),
+                    id2 = c("a", "b", "c", "e"),
+                    salary = c(60000, 55000, 70000, 80000),
+                    dept = c("IT", "Marketing", "Sales", "IT"))
R> join(df1, df2, on = c("id1", "id2"), how = "full", column = TRUE)
\end{Sinput}
\begin{Soutput}
full join: df1[id1, id2] 3/4 (75%) <m:m> df2[id1, id2] 3/4 (75%)
  id1 id2 name age salary      dept   .join
1   1   a John  35  60000        IT matched
2   1   b Jane  28     NA      <NA>     df1
3   2   b  Bob  42  55000 Marketing matched
4   3   c Carl  50  70000     Sales matched
5   3   e <NA>  NA  80000        IT     df2
\end{Soutput}
\end{Schunk}
An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}). Users can also invoke the \code{validate} argument to check the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
\begin{Schunk}
\begin{Sinput}
R> join(df1, df2, on = c("id1", "id2"), validate = "1:1", attr = "join") |>
+    attr("join") |> str(width = 70, strict = "cut")
\end{Sinput}
\begin{Soutput}
